Motor Learning Without Moving: Proprioceptive and Predictive Hand Localization After Passive Visuoproprioceptive Discrepancy Training

### 2018-11-28 paper/paper.tex
added files to compile the paper

---
title: "Motor Learning Without Moving: Proprioceptive and Predictive Hand Localization After Passive Visuoproprioceptive Discrepancy Training"
author:
  - name: Ahmed A. Mostafa
    affiliation: [YORK,MANSOURA]
#    affiliation: Faculty of Physical Education, Mansoura University, Mansoura, Egypt
  - name: Bernard Marius 't Hart
    affiliation: YORK
    email: thartbm@gmail.com
    corresponding: thartbm@gmail.com
  - name: Denise Y.P. Henriques
    email: deniseh@yorku.ca
    affiliation: YORK
address:
  - code: YORK
    address: CVR / Kinesiology and Health Science, York University, Toronto, Ontario, Canada
  - code: MANSOURA
    address: Faculty of Physical Education, Mansoura University, Mansoura, Egypt

abstract: |
  An accurate estimate of limb position is necessary for movement planning. Where we localize our unseen hand after a reach depends on felt hand position, or proprioception, but this is usually neglected in favour of predicted sensory consequences based on efference copies of motor commands. Both sources of information should contribute, so here we set out to further investigate how much of hand localization depends on proprioception and how much on predicted sensory consequences. We use a passive training paradigm with rotated visual feedback that eliminates the possibility to update predicted sensory consequences, but still recalibrates proprioception. After this training we measure participants’ hand location estimates based on both efference-based predictions and afferent proprioceptive signals with self-generated hand movements as well as based on proprioception only with robot-generated movements. We find indistinguishable shifts in hand localization after robot- and self-generated hand movements, and changes in open-loop reaches. Both motor and proprioceptive changes are only slightly smaller as those after training with self-generated movements, confirming that proprioception plays a large role in estimating limb position and in planning movements. (data: https://doi.org/10.17605/osf.io/zfdth, preprint: https://doi.org/10.1101/384941)

header-includes:
   - \usepackage{threeparttable}
   - \usepackage[table]{xcolor}
   - \usepackage{graphicx}
   - \usepackage{tabularx}
   - \usepackage{longtable}
   - \usepackage{siunitx}
bibliography: bibliography.bib
output: rticles::plos_article
csl: plos.csl
---

# Introduction

Sensory information is central to how we control all our movements. Our brain is even thought to use predicted sensory consequences derived from efferent copies of motor commands for motor control @Duhamel1992a. When training with rotated visual feedback of the hand, we update these predictions @Wolpert1995. Additionally, such training leads to a recalibration of our sense of felt hand position - “proprioception” - to be more aligned with the distorted visual feedback @Cressman2009. Both of these changes have been measured by asking people to localize where their unseen hand is – before and after training [@Clayton2013; @Izawa2012b; @Jones2012a; @Synofzik2008; @Yavari2016]. While our lab has previously found that proprioception accounts for a large portion of the change in hand localization @Hart2016, it is far from clear how much each process contributes or how to tease them apart. Here we use passive training, that removes the need to update predicted sensory consequences, in an attempt to isolate the contribution of proprioception to hand localization.

The predicted sensory consequences of movements may play several important roles in motor learning and control. Predicted sensory consequences allow us to correct our movements before sensory error signals are available, they can be used to select movements that best achieve our goals and they may inform us on the location of our limbs. Hence measuring predicted sensory consequences is valuable in movement research. In visuomotor rotation adaptation tasks, the actual sensory outcome is systematically different from the expected outcome, so that participants update their predictions on the outcome of the trained movements. In previous experiments, people were asked to make a movement and then indicate the location of, or “localize,” their unseen, right hand, before and after training with rotated visual feedback [@Izawa2012b; @Synofzik2008]. Since there was no visual information available to the participants, the predicted sensory consequences of the movement should be used in localizing the unseen hand. Both studies found a significant shift in hand localization, providing evidence that predicted sensory consequences are indeed updated as a result of visuomotor rotation adaptation.

However, our lab has shown that our sense of where we feel our hand to be, proprioception, is also reliably recalibrated after visuomotor rotation adaptation [@Cressman2009; @Hart2016; @Mostafa2014; @Mostafa2015; @Maksimovic2018; @Henriques2012; @Cressman2010a; @Barkley2014; @Salomonczyk2011; @Nourouzpour2014; @Henriques2014]. This has also been shown by other labs @Cameron2012 and a comparable proprioceptive change is induced with force-field adaptation @Ostry2016. As proprioception also informs us on the location of our limbs, we have on occasion used a task that is very similar to hand localization to investigate this [@Cressman2009; @Jones2012a; @Clayton2014; @Ruttle2018]. Although proprioceptive recalibration has been largely ignored as an explanation for changes in hand localization, we and others have shown that it accounts for a substantial part of the changes in localization, along with updates in predicted sensory consequences [@Hart2016; @Cameron2012]. Nevertheless, it is far from clear how much proprioception and prediction each contribute to hand localization.

Here we intend to further examine the contribution of proprioception to hand localization. To do this, we use passive training, where a robot arm moves the participant’s arm out, so that the cursor always directly goes to the target [@Cameron2012; @Cressman2010; @Salomonczyk2013]. This means there is no efference copy available and no visuomotor error-signal, both of which are required to update predicted sensory consequences. However, we impose a discrepancy between vision and proprioception that drives proprioceptive recalibration. Thus, changes in hand localization after this type of training should be due to proprioceptive recalibration only. We use the same experimental protocol as before @Hart2016, so that we can compare localization shifts between the two different types of training, and can better assess the contributions of predicted sensory consequences and proprioception to hand localization.

# Methods

@@

## Participants

Twenty-five right-handed participants were recruited for this study. One participant was excluded for not following task instructions, and three were excluded for low performance on a task that ensures attention during the passive training. All analyses presented here pertain to the remaining twenty-one participants (13 females and 8 males, mean age: 20.1 +/- 2.3 years), but the data of the three low-performing participants is included in the online \href{https://osf.io/zfdth}{dataset} @Mostafa2017. All had normal or corrected-to-normal vision, and provided prior, written informed consent in accordance with the ethical guidelines set by the York Human subjects Review Subcommittee and received credit toward an undergraduate psychology course. Participants were screened verbally, and all reported being right handed and not having any history of visual, neurological, and/or motor dysfunction.

![**Setup and tasks:** **a)** Participants moved their unseen right hand with visual feedback on hand position provided through a mirror (middle surface) half-way between their hand and the monitor (top surface). A touchscreen located just above the hand was used to collect responses for the localization tasks (bottom surface). **b)** Training task. The target, shown as a yellow disc, is located 10 cm away from the home position at 45°. In the rotated training tasks, the cursor (shown here as a green circle) represents the hand position rotated 30° relative to the home position. **c)** No-cursor reach task. Targets are located 10 cm away from the home position at 15°, 25°, 35°, 45°, 55°, 65°, and 75°, shown by the yellow circles here (only one was shown on each trial). While reaching to one of these targets, no visual feedback on hand position is provided. d) Localization task. The participants’ unseen, right hands moved out, and subsequently participants indicated the direction of the hand movement by indicating a location on an arc using a touch screen with their visible left index finger](Fig1_setup.eps)

## Setup

Participants sat in a height-adjustable chair to ensure that they could easily see and reach all targets presented on a reflective screen (see Fig 1). During all tasks, they held the vertical handle on a two-joint robot manipulandum (Interactive Motion Technologies Inc., Cambridge, MA) with their right hand so that their thumb rested on top of the handle. A monitor (Samsung 510 N, refresh rate 60 Hz) was mounted 11 cm above the reflective screen, such that images displayed on the monitor appeared to lie in the horizontal plane where the right hand was moving. The reflective screen was mounted horizontally 18 cm above the robot manipulandum. A touch screen was mounted 13 cm underneath the reflective surface, so that subjects could indicate the location of the unseen right-hand locations (specifically the unseen thumb) with their visible left hand, which was lit up with a small spot light (only in localization tasks). The room lights were dimmed and the participants’ view of their right hand was blocked by the reflective screen, as well as a dark cloth draped between the touch screen and participants’ right shoulders.

## Procedure

@@

$

@@

\begin{tablenotes}
\item 
\end{tablenotes}
\end{threeparttable}
\end{table}
$

## Exposure training

In what we call ‘exposure training’ the participants did not move their hand toward the target, but the robot did. In this task (Table 1), the right hand was represented by a cursor (green disk, 1 cm in diameter, Fig 1b) located directly above participant’s thumb. The robot moved the participant’s unseen right hand (and the cursor) along a direct path toward a visual yellow target disk and back to the starting position (1 cm in diameter, Fig 1b). The home position was located approximately 20 cm in front of participants and the visual target located 10 cm from the home position at 45° (Fig 1b). In order to make sure participants were paying attention to the cursor, the cursor was switched off for 2 screen refreshes (~33.3ms) on 50\% of the trials at a random distance between 4 and 9 cm from the home position and participants were asked to report this using a button press with the left hand. Performance on this task was used to screen participants.

During the first half of the experiment, the cursor and hand path were aligned during exposure training. In the second part of the experiment, the “rotated” session, the same visual training target at 45° was used, and the cursor kept moving straight to this target. However, the robot-generated hand path gradually rotated 30° CCW (Fig 1b) with respect to the visible target and the cursor in increments of 0.75°/trial, so that the full rotation was reached after 45 trials. This mimicked error-free responses to a gradual visuomotor rotation of 30° CW. The initial training consisted of 50 trials in the aligned part and 90 in the rotated part. In between open-loop reach tasks and localization tasks (Table 1) extra training tasks were done, each of which consisted of 10 trials in the aligned part of the experiment and 60 trials in the rotated part (Table 1).

## No-cursor reaching

@@

## Localization

In this task (Fig 1d; Table 1, dark gray rows) participants indicated where they thought their unseen right hand was after a movement. First, an arc appeared, spanning from 0° to 90° and located 10 cm away from the home position and the participants’ unseen, right hand moved out from the home position in a direction towards towards a point on the arc. The hand was stopped by the robot at 10 cm from the home position and then, to prevent the online proprioceptive signals from overriding the predictive signals [@Izawa2012b; @Hart2016], the hand was moved back to the home position using the same kind of constrained path as used for the return movements in the no-cursor task. Participants indicated with the index finger of their visible, left hand on the touch screen mounted directly above the robot handle where they thought their trained hand had crossed the arc.

Crucially, there were two variations of this task. First, in the ‘active’ localization task participants generated the movement themselves, as they could freely move their unseen right hand from the home position to any point on the arc. Second, there was a ‘passive’ localization task where the robot moved the participants’ hand out and back, to the same locations the participants moved to in the preceding ‘active’ localization task in a shuffled order (hence, active localization is done first). In active localization, participants have access to both proprioceptive information as well as an efference-based prediction of sensory consequences, but in passive localization, only proprioception should be available. The active and passive localization task each consisted of 25 trials, and each of the tasks was done a total of four times; twice after aligned and twice after rotated training.

## Classic training

The paradigm described above is an exact replica of a paradigm we used earlier @Hart2016 with two exceptions. First, we used exposure training here, instead of the standard reach training with volitional movements, which we will call ‘classic’ training. Second, all localization is delayed until the right hand has returned to the home position in this study (see Table 1), so that instead of both delayed and online localization we have two repetitions of each delayed localization task. With this paradigm we can compare changes in localization and no-cursor reaches change after exposure training with changes in the same measures after classic visuomotor adaptation training.

## Analyses

Prior to any analyses, both the localization responses and the no-cursor reach data were visually inspected and trials where the participants did not follow task instruction were removed (e.g. several movements back and forth, or a touch-screen response on the home position, instead of on the white arc).

## Localization

Localization responses were taken as the (signed) angular difference between vectors through the home position and the actual hand position as well as the location indicated on the touch screen. Prior to analyses, idiosyncratic differences in performing this task were countered. Before conversion to degrees angle, a circle with a 10 cm radius was fit to the touch screen responses of each participant and the offset of this circle’s centre was subtracted from all response coordinates, so that all responses fell close to the arc. Then, a smoothed spline was fit to every participant’s response errors in each of the four localization tasks (aligned vs. rotated and active vs. passive) and these were used to obtain localization errors at the same locations used for the no-cursor reaches (15°, 25°, 35°, 45°, 55°, 65° and 75°), but only if that location fell within the range of the data (i.e. we only interpolate). This way localization responses could be compared across participants despite the freely chosen reach directions. At the 15° location 7/21 participants didn’t have an estimate in one or more of the four localization tasks (in the “classic” data it was also 7/21). While that data is shown in the figures, we did not use it for analysis.

First we test if localization responses shifted following rotated exposure training compared to aligned. We then test if the shift in localization responses is different for active and passive localization, and we run analyses comparing localization after exposure training with localization after “classic” training. Finally, we explore the generalization of localization responses and if they are different between the groups doing classic and exposure training.

## Reach aftereffects

To assess any reach adaptation that may have occurred after exposure training we analyzed reach endpoint errors in no-cursor trials. Reach endpoint errors were the (signed) angular difference between a vector from the home position to reach endpoint and a vector from the home position to the target. We obtained reach aftereffects by subtracting reach endpoint errors after aligned training from those after rotated training. No-cursor endpoint errors were analyzed to see if participants adapted the direction of their reaching movements after rotated exposure training. We also tested if any such change decayed, i.e. if it was the same immediately after exposure training, or when a localization task was done in between exposure training and no-cursor reaches. Furthermore, we tested if the generalization of reach aftereffects is different between exposure and classic training and if there is any generalization of reach aftereffects.

Pre-processing and analyses were done in R 3.4.4 @RCoreTeam2017, using lme4, lmerTest and various other packages. Most analyses used linear mixed effects models, since there is some missing localization data. These were “converted” to more readable ANOVA-like output, using a Satterthwaite approximation @Luke2017. Highly similar results were obtained with a Chi-square approximation. Data, scripts and a notebook with analyses have been made available on the Open Science Framework @Mostafa2017 ([https://osf.io/zfdth](https://osf.io/zfdth/)).

In short, this experiment allowed us to test how mere exposure to a visual-proprioceptive discrepancy changes both reach aftereffects and hand localization responses, and compare them with those obtained after more regular, “classic” training.

# Results

In this study we intend to further elucidate the relative contributions of (updated) predicted sensory consequences and (recalibrated) proprioception to hand localization. We can parcel out these contributions by measuring hand localization after both robot-generated and self-generated movements. Finally we compare the data from the current experiment with those obtained in an earlier study that used an identical paradigm, but trained with self-generated movements, or “classic” training.

![**Hand localization:** The shifts of the angles of touchscreen responses in all variations of the localization task, using spline-interpolated estimates for hand angles matching the reach targets in the no-cursor reach block. **a)** Localization shifts after exposure training. Dark blue: active localization shifts, Light blue: passive localization shifts. **b)** Localization shifts after classic training. Dark red: active localization shifts, Orange: passive localization shifts. The dashed line segments illustrate that the 15° data is not used for statistical analyses (see Methods). **c)** Generalization curves of active localization shifts after exposure training (blue) and classic training (red). Shaded areas: 95\% confidence intervals for the peak of the generalization curve (red and blue lines through shaded area indicate 50\% points). Downward black arrow: visual trained target. Upward black arrow: hand location during training.](Fig2_plain.pdf)

## Localization

@@

To test if rotated exposure training induces changes in hand localization, we fit an LME to the localization errors throughout the workspace using session (aligned or rotated), movement type (active and passive) and hand angle (25°, 35°, 45°, 55°, 65° and 75°), and all interactions as fixed effects and participant as random effect. There was an effect of session (F(1,450.5)=155.8; p<0.001), showing that exposure training leads to changes in hand localization. There was also an effect of hand angle (F(5,450.6)=6.54; p<.001) and an interaction between hand angle and session (F(5,450.3)=8.25; p<.001), which we’ll explore below, but no other effects (all p>.60). Since localization responses did shift, we use the difference between hand localization after rotated training and after aligned training (as plotted in Fig 2) for further tests.

>>>>

% Template for PLoS
% Version 3.4 January 2017
\documentclass[10pt,letterpaper]{article}
\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}

% amsmath and amssymb packages, useful for mathematical formulas and symbols
\usepackage{amsmath,amssymb}

% Use adjustwidth environment to exceed column width (see example table in text)
\usepackage{changepage}

% Use Unicode characters when possible
\usepackage[utf8x]{inputenc}

% textcomp package and marvosym package for additional characters
\usepackage{textcomp,marvosym}

% cite package, to clean up citations in the main text. Do not remove.
% \usepackage{cite}

% Use nameref to cite supporting information files (see Supporting Information section for more info)
\usepackage{nameref,hyperref}

% line numbers
\usepackage[right]{lineno}

% ligatures disabled
\usepackage{microtype}
\DisableLigatures[f]{encoding = *, family = * }

% color can be used to apply background shading to table cells only
\usepackage[table]{xcolor}

% array package and thick rules for tables
\usepackage{array}

% create "+" rule type for thick vertical lines
\newcolumntype{+}{!{\vrule width 2pt}}

% create \thickcline for thick horizontal lines of variable length
\newlength\savedwidth
\newcommand\thickcline[1]{%
  \noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
  \cline{#1}%
  \noalign{\vskip\arrayrulewidth}%
  \noalign{\global\arrayrulewidth\savedwidth}%
}

% \thickhline command for thick horizontal lines that span the table
\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
\hline
\noalign{\global\arrayrulewidth\savedwidth}}

% Remove comment for double spacing
%\usepackage{setspace}
%\doublespacing

% Text layout
\raggedright
\setlength{\parindent}{0.5cm}
\textwidth 5.25in
\textheight 8.75in

% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
% Captions will be left justified
\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
\renewcommand{\figurename}{Fig}

% Use the PLoS provided BiBTeX style
% \bibliographystyle{plos2015}

% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

% Leave date blank
\date{}

% Header and Footer with logo
\usepackage{lastpage,fancyhdr,graphicx}
\usepackage{epstopdf}
\pagestyle{myheadings}
\pagestyle{fancy}
\fancyhf{}
\setlength{\headheight}{27.023pt}
\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
\rfoot{\thepage/\pageref{LastPage}}
\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
\fancyheadoffset[L]{2.25in}
\fancyfootoffset[L]{2.25in}
\lfoot{\sf PLOS}

%% Include all macros below
\newcommand{\lorem}{{\bf LOREM}}
\newcommand{\ipsum}{{\bf IPSUM}}

\usepackage{threeparttable}
\usepackage[table]{xcolor}
\usepackage{graphicx}
\usepackage{tabularx}
\usepackage{longtable}
\usepackage{siunitx}

\usepackage{forarray}
\usepackage{xstring}
\newcommand{\getIndex}[2]{
  \ForEach{,}{\IfEq{#1}{\thislevelitem}{\number\thislevelcount\ExitForEach}{}}{#2}
}

\setcounter{secnumdepth}{0}

\newcommand{\getAff}[1]{
  \getIndex{#1}{YORK,MANSOURA}
}

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\begin{document}
\vspace*{0.2in}

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Motor Learning Without Moving: Proprioceptive and Predictive Hand
Localization After Passive Visuoproprioceptive Discrepancy Training} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
\\
Ahmed A. Mostafa\textsuperscript{\getAff{YORK}, \getAff{MANSOURA}},
Bernard Marius 't Hart\textsuperscript{\getAff{YORK}}\textsuperscript{*},
Denise Y.P. Henriques\textsuperscript{\getAff{YORK}}\\
\bigskip
\textbf{\getAff{YORK}}CVR / Kinesiology and Health Science, York University, Toronto, Ontario,
Canada\\
\textbf{\getAff{MANSOURA}}Faculty of Physical Education, Mansoura University, Mansoura, Egypt\\
\bigskip
* Corresponding author: thartbm@gmail.com\\
\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
An accurate estimate of limb position is necessary for movement
planning. Where we localize our unseen hand after a reach depends on
felt hand position, or proprioception, but this is usually neglected in
favour of predicted sensory consequences based on efference copies of
motor commands. Both sources of information should contribute, so here
we set out to further investigate how much of hand localization depends
on proprioception and how much on predicted sensory consequences. We use
a passive training paradigm with rotated visual feedback that eliminates
the possibility to update predicted sensory consequences, but still
recalibrates proprioception. After this training we measure
participants' hand location estimates based on both efference-based
predictions and afferent proprioceptive signals with self-generated hand
movements as well as based on proprioception only with robot-generated
movements. We find indistinguishable shifts in hand localization after
robot- and self-generated hand movements, and changes in open-loop
reaches. Both motor and proprioceptive changes are only slightly smaller
as those after training with self-generated movements, confirming that
proprioception plays a large role in estimating limb position and in
planning movements. (data: https://doi.org/10.17605/osf.io/zfdth,
preprint: https://doi.org/10.1101/384941)

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step.
% Author Summary not valid for PLOS ONE submissions.

\linenumbers

% Use "Eq" instead of "Equation" for equation citations.
\section{Introduction}\label{introduction}

Sensory information is central to how we control all our movements. Our
brain is even thought to use predicted sensory consequences derived from
efferent copies of motor commands for motor control {[}1{]}. When
training with rotated visual feedback of the hand, we update these
predictions {[}2{]}. Additionally, such training leads to a
recalibration of our sense of felt hand position - ``proprioception'' -
to be more aligned with the distorted visual feedback {[}3{]}. Both of
these changes have been measured by asking people to localize where
their unseen hand is -- before and after training {[}4--8{]}. While our
lab has previously found that proprioception accounts for a large
portion of the change in hand localization {[}9{]}, it is far from clear
how much each process contributes or how to tease them apart. Here we
use passive training, that removes the need to update predicted sensory
consequences, in an attempt to isolate the contribution of
proprioception to hand localization.

The predicted sensory consequences of movements may play several
important roles in motor learning and control. Predicted sensory
consequences allow us to correct our movements before sensory error
signals are available, they can be used to select movements that best
achieve our goals and they may inform us on the location of our limbs.
Hence measuring predicted sensory consequences is valuable in movement
research. In visuomotor rotation adaptation tasks, the actual sensory
outcome is systematically different from the expected outcome, so that
participants update their predictions on the outcome of the trained
movements. In previous experiments, people were asked to make a movement
and then indicate the location of, or ``localize,'' their unseen, right
hand, before and after training with rotated visual feedback {[}5,7{]}.
Since there was no visual information available to the participants, the
predicted sensory consequences of the movement should be used in
localizing the unseen hand. Both studies found a significant shift in
hand localization, providing evidence that predicted sensory
consequences are indeed updated as a result of visuomotor rotation
adaptation.

However, our lab has shown that our sense of where we feel our hand to
be, proprioception, is also reliably recalibrated after visuomotor
rotation adaptation {[}3,9--18{]}. This has also been shown by other
labs {[}19{]} and a comparable proprioceptive change is induced with
force-field adaptation {[}20{]}. As proprioception also informs us on
the location of our limbs, we have on occasion used a task that is very
similar to hand localization to investigate this {[}3,6,21,22{]}.
Although proprioceptive recalibration has been largely ignored as an
explanation for changes in hand localization, we and others have shown
that it accounts for a substantial part of the changes in localization,
along with updates in predicted sensory consequences {[}9,19{]}.
Nevertheless, it is far from clear how much proprioception and
prediction each contribute to hand localization.

Here we intend to further examine the contribution of proprioception to
hand localization. To do this, we use passive training, where a robot
arm moves the participant's arm out, so that the cursor always directly
goes to the target {[}19,23,24{]}. This means there is no efference copy
available and no visuomotor error-signal, both of which are required to
update predicted sensory consequences. However, we impose a discrepancy
between vision and proprioception that drives proprioceptive
recalibration. Thus, changes in hand localization after this type of
training should be due to proprioceptive recalibration only. We use the
same experimental protocol as before {[}9{]}, so that we can compare
localization shifts between the two different types of training, and can
better assess the contributions of predicted sensory consequences and
proprioception to hand localization.

\section{Methods}\label{methods}

@@

\subsection{Participants}\label{participants}

Twenty-five right-handed participants were recruited for this study. One
participant was excluded for not following task instructions, and three
were excluded for low performance on a task that ensures attention
during the passive training. All analyses presented here pertain to the
remaining twenty-one participants (13 females and 8 males, mean age:
20.1 +/- 2.3 years), but the data of the three low-performing
participants is included in the online
\href{https://osf.io/zfdth}{dataset} {[}25{]}. All had normal or
corrected-to-normal vision, and provided prior, written informed consent
in accordance with the ethical guidelines set by the York Human subjects
Review Subcommittee and received credit toward an undergraduate
psychology course. Participants were screened verbally, and all reported
being right handed and not having any history of visual, neurological,
and/or motor dysfunction.

\begin{figure}
\centering
\includegraphics{Fig1_setup.eps}
\caption{\textbf{Setup and tasks:} \textbf{a)} Participants moved their
unseen right hand with visual feedback on hand position provided through
a mirror (middle surface) half-way between their hand and the monitor
(top surface). A touchscreen located just above the hand was used to
collect responses for the localization tasks (bottom surface).
\textbf{b)} Training task. The target, shown as a yellow disc, is
located 10 cm away from the home position at 45°. In the rotated
training tasks, the cursor (shown here as a green circle) represents the
hand position rotated 30° relative to the home position. \textbf{c)}
No-cursor reach task. Targets are located 10 cm away from the home
position at 15°, 25°, 35°, 45°, 55°, 65°, and 75°, shown by the yellow
circles here (only one was shown on each trial). While reaching to one
of these targets, no visual feedback on hand position is provided. d)
Localization task. The participants' unseen, right hands moved out, and
subsequently participants indicated the direction of the hand movement
by indicating a location on an arc using a touch screen with their
visible left index finger}
\end{figure}

\subsection{Setup}\label{setup}

Participants sat in a height-adjustable chair to ensure that they could
easily see and reach all targets presented on a reflective screen (see
Fig 1). During all tasks, they held the vertical handle on a two-joint
robot manipulandum (Interactive Motion Technologies Inc., Cambridge, MA)
with their right hand so that their thumb rested on top of the handle. A
monitor (Samsung 510 N, refresh rate 60 Hz) was mounted 11 cm above the
reflective screen, such that images displayed on the monitor appeared to
lie in the horizontal plane where the right hand was moving. The
reflective screen was mounted horizontally 18 cm above the robot
manipulandum. A touch screen was mounted 13 cm underneath the reflective
surface, so that subjects could indicate the location of the unseen
right-hand locations (specifically the unseen thumb) with their visible
left hand, which was lit up with a small spot light (only in
localization tasks). The room lights were dimmed and the participants'
view of their right hand was blocked by the reflective screen, as well
as a dark cloth draped between the touch screen and participants' right
shoulders.

\subsection{Procedure}\label{procedure}

@@

\$

@@

\begin{tablenotes}
\item 
\end{tablenotes}
\end{threeparttable}
\end{table}

\$

\subsection{Exposure training}\label{exposure-training}

In what we call `exposure training' the participants did not move their
hand toward the target, but the robot did. In this task (Table 1), the
right hand was represented by a cursor (green disk, 1 cm in diameter,
Fig 1b) located directly above participant's thumb. The robot moved the
participant's unseen right hand (and the cursor) along a direct path
toward a visual yellow target disk and back to the starting position (1
cm in diameter, Fig 1b). The home position was located approximately 20
cm in front of participants and the visual target located 10 cm from the
home position at 45° (Fig 1b). In order to make sure participants were
paying attention to the cursor, the cursor was switched off for 2 screen
refreshes (\textasciitilde{}33.3ms) on 50\% of the trials at a random
distance between 4 and 9 cm from the home position and participants were
asked to report this using a button press with the left hand.
Performance on this task was used to screen participants.

During the first half of the experiment, the cursor and hand path were
aligned during exposure training. In the second part of the experiment,
the ``rotated'' session, the same visual training target at 45° was
used, and the cursor kept moving straight to this target. However, the
robot-generated hand path gradually rotated 30° CCW (Fig 1b) with
respect to the visible target and the cursor in increments of
0.75°/trial, so that the full rotation was reached after 45 trials. This
mimicked error-free responses to a gradual visuomotor rotation of 30°
CW. The initial training consisted of 50 trials in the aligned part and
90 in the rotated part. In between open-loop reach tasks and
localization tasks (Table 1) extra training tasks were done, each of
which consisted of 10 trials in the aligned part of the experiment and
60 trials in the rotated part (Table 1).

\subsection{No-cursor reaching}\label{no-cursor-reaching}

@@

\subsection{Localization}\label{localization}

In this task (Fig 1d; Table 1, dark gray rows) participants indicated
where they thought their unseen right hand was after a movement. First,
an arc appeared, spanning from 0° to 90° and located 10 cm away from the
home position and the participants' unseen, right hand moved out from
the home position in a direction towards towards a point on the arc. The
hand was stopped by the robot at 10 cm from the home position and then,
to prevent the online proprioceptive signals from overriding the
predictive signals {[}5,9{]}, the hand was moved back to the home
position using the same kind of constrained path as used for the return
movements in the no-cursor task. Participants indicated with the index
finger of their visible, left hand on the touch screen mounted directly
above the robot handle where they thought their trained hand had crossed
the arc.

Crucially, there were two variations of this task. First, in the
`active' localization task participants generated the movement
themselves, as they could freely move their unseen right hand from the
home position to any point on the arc. Second, there was a `passive'
localization task where the robot moved the participants' hand out and
back, to the same locations the participants moved to in the preceding
`active' localization task in a shuffled order (hence, active
localization is done first). In active localization, participants have
access to both proprioceptive information as well as an efference-based
prediction of sensory consequences, but in passive localization, only
proprioception should be available. The active and passive localization
task each consisted of 25 trials, and each of the tasks was done a total
of four times; twice after aligned and twice after rotated training.

\subsection{Classic training}\label{classic-training}

The paradigm described above is an exact replica of a paradigm we used
earlier {[}9{]} with two exceptions. First, we used exposure training
here, instead of the standard reach training with volitional movements,
which we will call `classic' training. Second, all localization is
delayed until the right hand has returned to the home position in this
study (see Table 1), so that instead of both delayed and online
localization we have two repetitions of each delayed localization task.
With this paradigm we can compare changes in localization and no-cursor
reaches change after exposure training with changes in the same measures
after classic visuomotor adaptation training.

\subsection{Analyses}\label{analyses}

Prior to any analyses, both the localization responses and the no-cursor
reach data were visually inspected and trials where the participants did
not follow task instruction were removed (e.g.~several movements back
and forth, or a touch-screen response on the home position, instead of
on the white arc).

\subsection{Localization}\label{localization-1}

Localization responses were taken as the (signed) angular difference
between vectors through the home position and the actual hand position
as well as the location indicated on the touch screen. Prior to
analyses, idiosyncratic differences in performing this task were
countered. Before conversion to degrees angle, a circle with a 10 cm
radius was fit to the touch screen responses of each participant and the
offset of this circle's centre was subtracted from all response
coordinates, so that all responses fell close to the arc. Then, a
smoothed spline was fit to every participant's response errors in each
of the four localization tasks (aligned vs.~rotated and active
vs.~passive) and these were used to obtain localization errors at the
same locations used for the no-cursor reaches (15°, 25°, 35°, 45°, 55°,
65° and 75°), but only if that location fell within the range of the
data (i.e.~we only interpolate). This way localization responses could
be compared across participants despite the freely chosen reach
directions. At the 15° location 7/21 participants didn't have an
estimate in one or more of the four localization tasks (in the
``classic'' data it was also 7/21). While that data is shown in the
figures, we did not use it for analysis.

First we test if localization responses shifted following rotated
exposure training compared to aligned. We then test if the shift in
localization responses is different for active and passive localization,
and we run analyses comparing localization after exposure training with
localization after ``classic'' training. Finally, we explore the
generalization of localization responses and if they are different
between the groups doing classic and exposure training.

\subsection{Reach aftereffects}\label{reach-aftereffects}

To assess any reach adaptation that may have occurred after exposure
training we analyzed reach endpoint errors in no-cursor trials. Reach
endpoint errors were the (signed) angular difference between a vector
from the home position to reach endpoint and a vector from the home
position to the target. We obtained reach aftereffects by subtracting
reach endpoint errors after aligned training from those after rotated
training. No-cursor endpoint errors were analyzed to see if participants
adapted the direction of their reaching movements after rotated exposure
training. We also tested if any such change decayed, i.e.~if it was the
same immediately after exposure training, or when a localization task
was done in between exposure training and no-cursor reaches.
Furthermore, we tested if the generalization of reach aftereffects is
different between exposure and classic training and if there is any
generalization of reach aftereffects.

Pre-processing and analyses were done in R 3.4.4 {[}26{]}, using lme4,
lmerTest and various other packages. Most analyses used linear mixed
effects models, since there is some missing localization data. These
were ``converted'' to more readable ANOVA-like output, using a
Satterthwaite approximation {[}27{]}. Highly similar results were
obtained with a Chi-square approximation. Data, scripts and a notebook
with analyses have been made available on the Open Science Framework
{[}25{]} (\href{https://osf.io/zfdth/}{https://osf.io/zfdth}).

In short, this experiment allowed us to test how mere exposure to a
visual-proprioceptive discrepancy changes both reach aftereffects and
hand localization responses, and compare them with those obtained after
more regular, ``classic'' training.

\section{Results}\label{results}

In this study we intend to further elucidate the relative contributions
of (updated) predicted sensory consequences and (recalibrated)
proprioception to hand localization. We can parcel out these
contributions by measuring hand localization after both robot-generated
and self-generated movements. Finally we compare the data from the
current experiment with those obtained in an earlier study that used an
identical paradigm, but trained with self-generated movements, or
``classic'' training.

\begin{figure}
\centering
\includegraphics{Fig2_plain.pdf}
\caption{\textbf{Hand localization:} The shifts of the angles of
touchscreen responses in all variations of the localization task, using
spline-interpolated estimates for hand angles matching the reach targets
in the no-cursor reach block. \textbf{a)} Localization shifts after
exposure training. Dark blue: active localization shifts, Light blue:
passive localization shifts. \textbf{b)} Localization shifts after
classic training. Dark red: active localization shifts, Orange: passive
localization shifts. The dashed line segments illustrate that the 15°
data is not used for statistical analyses (see Methods). \textbf{c)}
Generalization curves of active localization shifts after exposure
training (blue) and classic training (red). Shaded areas: 95\%
confidence intervals for the peak of the generalization curve (red and
blue lines through shaded area indicate 50\% points). Downward black
arrow: visual trained target. Upward black arrow: hand location during
training.}
\end{figure}

\subsection{Localization}\label{localization-2}

@@

To test if rotated exposure training induces changes in hand
localization, we fit an LME to the localization errors throughout the
workspace using session (aligned or rotated), movement type (active and
passive) and hand angle (25°, 35°, 45°, 55°, 65° and 75°), and all
interactions as fixed effects and participant as random effect. There
was an effect of session (F(1,450.5)=155.8; p\textless{}0.001), showing
that exposure training leads to changes in hand localization. There was
also an effect of hand angle (F(5,450.6)=6.54; p\textless{}.001) and an
interaction between hand angle and session (F(5,450.3)=8.25;
p\textless{}.001), which we'll explore below, but no other effects (all
p\textgreater{}.60). Since localization responses did shift, we use the
difference between hand localization after rotated training and after
aligned training (as plotted in Fig 2) for further tests.
